Анализ свойств вероятностных моделей в задачах обучения с экспертом

### 2022-01-30 paper/Grabovoy2022ExpertLearningAIT/Manuscript.tex
Updates from Overleaf

\documentclass[12pt]{a&t}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{bm}
\usepackage{url}
\usepackage{multirow}
\renewcommand{\baselinestretch}{2}
\usepackage{cmap}
\usepackage{autonum}
\usepackage{cite}

@@

\begin{document}  %%%!!!

@@

\title{Analysis of the properties of probabilistic models in learning problems with an expert}%
\thanks{The reported study was funded by Russian Foundation for Basic Research according to the research projects 20-37-90050, 19-07-01155 and National Technological Initiative project 13/1251/2018.}

\authors{А.В.~ГРАБОВОЙ (grabovoy.av@phystech.edu)\\
(Московский физико-технический институт),\\
В.В.~СТРИЖОВ, д-р~физ.-мат.~наук (strijov@phystech.edu)\\
(Вычислительный центр им. А.А. Дородницына ФИЦ ИУ РАН)}

@@

\begin{abstract}
This paper is devoted to the construction of interpretable models in machine learning. 
It solves the problem of approximating a given set of figures on a contour image.
Assumptions are introduced that the figures are curves of the second order.
When approximating figures, information about the type, location and form of curves, as well as the set of their possible transformations, is used.
Such information is called \textit{expert}, and the machine learning method based on expert information is called \textit {expert learning}.
It is assumed that the set of shapes is approximated by a set of \textit {local models}.
Each local model, based on expert information, approximates one shape in the contour image.
To build models, it is proposed to map second-order curves into a feature space where each local model is a linear model.
Thus, higher-order curves are recognized using the composition of linear models.
As an applied problem, the problem of approximating the iris of the eye on a contour image is considered.

\smallskip\\
\textit{Ключевые слова}: Mixture of experts; Expert learning;  Linear models; Interpretabe models.
\end{abstract}

\section{Introduction}
\label{intro}
Interpretable model building in machine learning~(Ribeiro et al. 2016) is one of the key challenges.
Modern solutions of the image classification problem based on deep learning networks ResNet, VGG, Intercept~(Kaiming et al. 2016) are poorly interpreted models.
The papers~(Han et al. 2020; Akhtar et al. 2018) show that deep learning networks are sensitive even to small noise in the data, which is due to their uninterpretability.

In this paper, we propose a \textit {training with an expert} method.
This method assumes the use of subject knowledge of experts to improve the quality of approximation, as well as to obtain interpretable machine learning models.
The subject knowledge of experts about the sample will be called~\textit {expert information}.
It is assumed that the use of expert information allows the sample to be approximated by simple interpretable models, such as linear models. Machine learning methods that take expert knowledge into account when building models are called~\textit {expert learning}.

This paper solves the problem of approximating second-order curves on a contour image. Second-order curves are selected for analysis, since they are easily described by linear models. In this case, these figures need to be restored in such applied problems as the problem of recognizing the iris of the eye~(Matveev 2010; Matveev et al. 2014; Bowyer et al. 2010), the problem of describing the particle track in the hadron collider~(Salamani et al. 2018). Expert information about a second-order curve allows you to map points on a plane into a new feature description, where each curve is approximated by one linear model. A model that approximates one curve is called a \textit {local model}. To approximate the entire contour image, it is required to approximate several second-order curves using several local models. In this paper, the following restrictions on images are introduced: a) the image consists only of second-order curves; b) the image is approximated by a small number of second-order curves; c) the number and type of curves in the image is known.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{explanation}
  \caption {Example: a) expert information of the first expert; b) baseline data; c) expert information of the second expert}
  \label{intro:fig2}
\end{figure}

Figure~\ref{intro:fig2} shows an example of second-order curves, as well as expert information on curves. Figure~\ref{intro:fig2} a shows the expert information of the first expert. Using this information, the first curve is fitted with a linear model and the second curve is noise. Figure~\ref{intro:fig2} b shows the expert information of the second expert. Using this information, the second curve is fitted with a linear model and the first curve is noise.

When approximating several curves on one contour image, a multi-model is built. An example of multi-models is a random forest~(Chen et al. 2012), tree boosting~(Chen et al. 2016), a mixture of experts~(Yuksel et al. 2012). In this paper, a mixture of experts is considered as a multi-model. Expert mixture is a multi-model that linearly weights local models that approximate a portion of the sample. The values of the weighting coefficients depend on the object for which the prediction is made. To solve the problem of a mixture of experts, a variational EM-algorithm~(Dempster et al. 1997; Bishop 2010; Peng et al. 1996) is used. The mixture of experts has many uses in a number of applications. In the paper~(Estabrooks et al. 2001), the text classification problem is solved. In the papers~(Cheung et al. 1995; Weigend et al. 2000; Cao 2003; Mossavat et al. 2010; Sminchisescu C et al. 2007; Tuerk 2001; Yumlu et al. 2003), a mixture of experts is used to predict time series for speech recognition, daily human activity, and prediction of the value of securities. In the paper~(Ebrahimpour et al. 2009), a mixture of experts was considered to solve the problem of recognizing handwritten numbers in images.

As an example, the problem of approximation of the iris image is considered. Figure~\ref{intro:fig1:real} shows an example of the image that needs to be approximated. In this paper, we consider a processed image, which is given in outline form, an example of such an image is shown in Figure~\ref{intro:fig1:outer}. Figure~\ref{intro:fig1:outer} shows two local circle models that approximate the iris of the eye. Circumferences are a simple example of a second order curve.

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{real_image}\label{intro:fig1:real}
\includegraphics[width=0.5\textwidth]{outline_image}\label{intro:fig1:outer}
\caption{An example of the image of the iris of the eye and its outline representation: a) the image of the iris of the eye; b) contour image of the iris and approximating the given image of the circumferences}
\label{intro:fig1}
\end{figure}

For the problem of approximating the iris of the eye, the following expert information is used: the iris of the eye is approximated by two concentric circumferences. Expert information is used to construct a feature description of plane points, as well as to build an optimization function. The part of the error function for optimization that uses expert information is called a regularizer. Thus, the information that the image of the circumferences is specified by the feature description, and the information that the concentric circumferences are specified using a special regularizer.

In a computational experiment, the quality of the approximation of the contour image is analyzed depending on the specified expert information and on the noise level in the synthetically generated data. The analysis of the quality of the approximation of the iris is carried out, depending on the amount of expert information that was used to build the model. Note that each approximated image is a separate set of points that need to be approximated.

\section{Statement of the problem of finding the parameters of the second-order curves in the image}
\label{sec:1}
Binary image is set:
$$ \mathbf{M} \in \{0, 1 \}^{m_1\times m_2},$$
where 1 corresponds to the black point of the image, and 0 corresponds to the white point of the background.
From the image $\mathbf {M} $, a sample$ \mathbf{C}$ is constructed, the elements of which are the coordinates $(x_i, y_i)$ of black points: $$\mathbf{C} \in \mathbb{R}^{N \times 2}. $$
The expert assumes that the image consists of a second-order curve~$\Omega$.
Let for a set of points $\mathbf {C} \in \mathbb {R}^{N \times 2} $ that form a curve $\Omega, $ expert information about the figure $E(\Omega) $is given.
The set $E (\Omega)$ consists of the shape $\Omega$ expected by the expert and the set of its admissible transformations. Based on the expert description, let us introduce mappings into a new problem for approximation:
\begin{equation}\label{eq1}
	K_{x}\bigl(E(\Omega)\bigr): \mathbb{R}^{2} \rightarrow \mathbb{R}^{n}, \quad K_{y}\bigl(E(\Omega)\bigr): \mathbb{R}^{2} \rightarrow \mathbb{R},
\end{equation} 
where~$K_{x}$ mapping objects to the attribute description of objects,~$n$is the number of features, and~$K_ {y}$ is a mapping to a target variable for an object. Applying the mappings~$K_ {x}, K_{y}$for the sample~$\mathbf {C}$ element by element we obtain:
\begin{equation}
\label{eq2}
	K_{x}\bigl(E(\Omega\bigr), \mathbf{c}) = \mathbf{x}, \quad  K_{y}\bigl(E(\Omega), \mathbf{c}\bigr) = y,
\end{equation}
where~$\mathbf{c} = (x_i, y_i)$ is a sample point $\mathbf{C}$.

Applying the mappings \eqref{eq2} to the original set of points $\mathbf {C} $, we obtain the sample
\begin{equation}
\label{eq4}
    \mathfrak{D} = \{(\mathbf{x}, y) \; | \; \forall \mathbf{c} \in \mathbf{C} \; \mathbf{x} = K_x(\mathbf{c}), \; y = K_y(\mathbf{c}) \}.
\end{equation}

We get that the original problem of curve approximation~$\Omega$ is reduced to approximation of the sample~$\mathfrak {D}$. In this paper, it is assumed that the sample~$\mathfrak {D}$ is approximated by a linear model:
\begin{equation}
	g(\mathbf{x}, \mathbf{w}) = \mathbf{x}^\mathsf{T} \mathbf{w},
\end{equation} 
where~$\mathbf{w}$ vector, the parameter to be found.

To find the optimal vector of parameters~$\hat {\mathbf {w}}$, it is required to solve the following optimization problem:
\begin{equation}
	\hat{\mathbf{w}} = \arg\min_{\mathbf{w}\in\mathbb{R}^n} \sum_{\left(\mathbf{x}, y\right) \in \mathfrak{D}}\|g(\mathbf{x}, \mathbf{w}) - y \|_2^2.
\end{equation}

Thus, the problem of approximating the original curve~$\Omega$ is reduced to solving the problem of linear regression, i.e. finding the components of the vector $\hat{\mathbf {w}}$ connecting the resulting $\mathbf {x}$ and $y$.

In the case when on the image $K$ the second-order curves  $\Omega_1, \dots, \Omega_K$, for each of which there is expert information $E_k = E (\Omega_k), \, k \in \{1, \dots , K\}$, the problem of constructing a multi-model called a mixture of $K$ experts is posed.
\begin{definition}
We call the multimodel $ f $ a mixture of K experts
\begin{equation}
	f = \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V})g_k(\mathbf{w}_k),  \quad \pi_k(\mathbf{x}, \mathbf{V}): \mathbb{R}^{n\times |\mathbf{V}|} \rightarrow [0, \, 1], \quad \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V}) = 1, 
\end{equation}
where $g_k$ is a local model called by the expert, $\mathbf{x}$is an attribute description of an object, $\pi_k$is a gateway function, the vector $\mathbf{w}_k$ are local model parameters, the vector $\mathbf{V}$ are gateway function parameters. In this paper, $g_k$ is a linear model.
\end{definition}

For each second-order curve, mappings (\ref{eq1}) are given. For convenience, we introduce the following notation: $K_x^k\bigr(\mathbf{c}\bigr) = K_x\bigr(\Omega_k, \mathbf{c}\bigr)$ and $K_y^k\bigr(\mathbf{c}\bigr) = K_y\bigr(\Omega_k, \mathbf{c}\bigr)$. Then, using local linear models, we construct a universal multi-model describing the curves $\Omega_1, \dots, \Omega_K$ on the image $\mathbf{M}$:
\begin{equation}
\label{5}
	f = \sum\limits_{\mathbf{c} \in \mathbf{C}} \sum_{k = 1}^{K} \pi_k(\mathbf{c}, \mathbf{V})g_k(K^k_{x}\bigl(\mathbf{c}), \mathbf{w}_k), 
\end{equation}
where $\pi_k$ is the  gateway function. In this paper, we consider a simple case, where~$\mathbf{x}=K^1_{x}\bigl(\mathbf{c})=\cdots=K^K_{x}\bigl(\mathbf{c}),$ then the expression~\eqref{5} is rewritten in the following simple form:
\begin{equation}
\label{5_1}
	f = \sum\limits_{\mathbf{c} \in \mathbf{C}} \sum_{k = 1}^{K} \pi_k(\mathbf{x}, \mathbf{V})g_k(\mathbf{x}, \mathbf{w}_k), 
\end{equation}
where the gateway function $\pi_k$ has the following form:
\begin{equation}
\label{6}
	\pi_k(\mathbf{x}, \mathbf{V}): \mathbb{R}^{n\times |\mathbf{V}|} \rightarrow [0, \, 1], \; \; \; \; \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V}) = 1,
\end{equation}
where $\mathbf{V}$ are the gateway function parameters, and $g_k$ is a local model.

In this paper
\begin{equation}
    \boldsymbol{\pi}(\mathbf{x}, \mathbf{V}) = \text{softmax}\bigl(\mathbf{V}_1^{\mathsf{T}}\boldsymbol{\sigma}(\mathbf{V}_2^{\mathsf{T}}\mathbf{x}) \bigr),
\end{equation}
where $\mathbf{V} = \{ \mathbf{V}_1, \, \mathbf{V}_2\}$ are the gateway function parameters,
$\mathbf{V}_1 \in \mathbb{R}^{p \times k}, \, \mathbf{V}_2 \in \mathbb{R}^{n \times p}$.

To find the optimal parameters of the multi-model, it is necessary to solve the following optimization problem:
\begin{equation}\label{9}
\mathcal{L} = \sum\limits_{(\mathbf{x}, y) \in \mathfrak{D}} \sum\limits_{k = 1}^{K} \pi_k(\mathbf{x}, \mathbf{V})(y - \mathbf{w}_k^{\mathsf{T}}\mathbf{x})^2 + R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr) \rightarrow \min_{\mathbf{V}, \mathbf{W}},
\end{equation}
where $\mathbf{W} = [\mathbf{w}_1, \dots, \mathbf{w}_k]$ are  parameters of local models, $R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr)$ is regularization parameters, based on expert information.

\section{Building an attribute description of figures}
\label{sec:3}
\paragraph{ Unified space for second-order curves.} An arbitrary second-order curve, the main axis of which is not parallel to the ordinate axis, is given by the following expression:
\[
\label{st:coef}
x^2 = B'xy+C'y^2+D'x+E'y+F',
\]
where the coefficients $B ', C'$ are subject to restrictions that depend on the type of the curve. The expression~\eqref{eq2} takes the following form:
\[
\label{st:K_map}
K_x\bigr(\mathbf{c}_i\bigr)=\left[x_iy_i, y_i^2, x_i, y_i, 1\right], \quad K_y\bigr(\mathbf{c}_i\bigr)=x_i^2,
\]
whence we obtain the linear regression problem for recovering the parameters~$ B ', C', D ', E', F '$ from the selected sample.

\paragraph{ Circumference.} As a special case of a second-order curve, we consider the circumference.
Let $(x_0, y_0)$ be the center of the circumference to be found on the binary image $\mathbf {M} $, and $r$ be its radius.
The sample elements $(x_i, y_i) \in \mathbf {C}$ are the locus of points, which is approximated by the equation of the circumference:
\begin{equation}
(x_i - x_0)^2 + (y_i - y_0)^2 = r^2.
\end{equation}
Expanding the brackets, we get:
\begin{equation}(2x_0)\cdot x_i + (2y_0)\cdot y_i + (r^2 - x_0^2 - y_0^2)\cdot 1 = x_i^2 + y_i^2 . 
\end{equation}
Then the presentations (\ref{eq2}) take the following form:
\begin{equation}
\label{10}
K_{x}(\mathbf{c}_i) = [x_i, \, y_i, \, 1] = \mathbf{x}, \,  K_{y}(\mathbf{c}_i) = x_i^2+y_i^2 = y.
\end{equation} 
Assign the linear regression problem \eqref{eq4}.
Vector components $\mathbf{w} = [w_0, \, w_1, \, w_2]^\mathsf{T}$, binding $\mathbf{x}$ and $y$, restore the parameters of the circumference: \begin{equation} x_0 = \frac{w_0}{2}, \; y_0 = \frac{w_1}{2}, \; r = \sqrt{w_3 + x_0^2 + y_0 ^2}.\end{equation}

\section{Composition of figures}
\label{sec:4}
o construct a composition of figures, we will use the expression~\eqref{9}, which takes the following form:
\begin{equation} 
\label{statment:optim:task}
\begin{aligned}
\mathcal{L} = \sum\limits_{\mathbf{c} \in \mathbf{C}} \sum\limits_{k = 1}^{K} \pi_k(\mathbf{c}, \mathbf{V})\left(K^{k}_y\bigr(\mathbf{c}\bigr) - \mathbf{w}_k^{\mathsf{T}}K^{k}_x\bigr(\mathbf{c}\bigr)\right)^2 + R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr) \rightarrow \min_{\mathbf{V}, \mathbf{W}},
\end{aligned}
\end{equation} 
where~$K^{k}_x, K^{k}_y$ expert representation of the~$k$-th expert. Assuming that all curves in the image are described by one attribute description $\mathbf {x} =K^{1}_x\bigr(\mathbf{c}\bigr)=\cdots=K^{K}_x\bigr(\mathbf{c}\bigr), x= K^{1}_y\bigr(\mathbf{c}\bigr)=\cdots=K^{K}_y\bigr(\mathbf{c}\bigr),$ we get the following optimization problem:
\begin{equation} 
\label{statment:optim:task:simp}
\begin{aligned}
\mathcal{L} = \sum\limits_{\left(\mathbf{x}, y\right) \in \mathfrak{D}} \sum\limits_{k = 1}^{K} \pi_k(\mathbf{x}, \mathbf{V})\left(y - \mathbf{w}_k^{\mathsf{T}}\mathbf{x}\right)^2 + R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr) \rightarrow \min_{\mathbf{V}, \mathbf{W}},
\end{aligned}
\end{equation}

>>>>

\documentclass[12pt]{a&t}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{bm}
\usepackage{url}
\usepackage{multirow}
\usepackage{cmap}
\usepackage{autonum}
\usepackage{cite}

@@

\begin{document}

@@

\title{Анализ свойств вероятностных моделей в задачах обучения с экспертом}%
\thanks{Исследование выполнено при финансовой поддержке Российского Фонда Фундаментальных Исследований в рамках научных проектов 20-37-90050, 19-07-01155 и проекта Национальной Технологической Инициативы 13/1251/2018.}

\authors{А.И.~БАЗАРОВА (bazarova.ai@phystech.edu)\\
(Московский физико-технический институт),\\
А.В.~ГРАБОВОЙ (grabovoy.av@phystech.edu)\\
(Московский физико-технический институт),\\
В.В.~СТРИЖОВ, д-р~физ.-мат.~наук (strijov@phystech.edu)\\
(Вычислительный центр им. А.А. Дородницына ФИЦ ИУ РАН)}

@@

\begin{abstract}
Данная работа посвящена построению интерпретируемых моделей машинного обучения.
Решается задача аппроксимации набора фигур на контурном изображении.
Вводятся предположения, что фигуры являются кривыми второго порядка.
При аппроксимации фигур используются информация о типе, расположении и форме кривых, а также о множестве их возможных преобразований.
Такая информация называется \textit{экспертной}, а метод машинного обучения, основанный на экспертной информации, называется \textit{обучение с экспертом}.
Предполагается, что набор фигур аппроксимируется набором \textit{локальных моделей}.
Каждая локальная модель, основанная на экспертной информации, аппроксимирует одну фигуру на контурном изображении.
Для построения моделей предлагается отображать кривые второго порядка в пространство признаков, в которой каждая локальная модель является линейной.
Таким образом, кривые второго порядка аппроксимируются набором линейных моделей.
В вычислительном эксперименте рассматривается задача аппроксимации радужной оболочки глаза на контурном изображении.

\smallskip\\
\textit{Ключевые слова}: смесь экспертов; экспертное обучение; линейные модели; интерпретируемые модели.
\end{abstract}

\section{Введение}
Современные решения задачи классификации изображений на основе сетей глубокого обучения ResNet, VGG, Intercept~\cite{Kaiming2015} представляют собой плохо интерпретируемые модели~\cite{Ribeiro2016}.
В~\cite{Han2020, Akhtar2018} показано, что сети глубокого обучения являются не устойчивыми даже к небольшому шуму в данных.

В данной работе предлагается метод \textit{обучение с экспертом}.
Метод предполагает использование экспертной информации для повышения качества аппроксимации, а также для получения интерпретируемых моделей машинного обучения.
Предметные знания экспертов о данных называются~\textit{экспертная информация}.
Предполагается, что использование экспертной информации позволяет аппроксимировать выборку простыми интерпретируемыми моделями, такими как линейные модели. Методы машинного обучения, учитывающие экспертные знания при построении моделей, называются~\textit{обучением с экспертом}.
В данной работе решается задача аппроксимации кривых второго порядка на контурном изображении.
В работе анализируются кривые второго порядка, так как они описываются линейными моделями. Параметры кривых второго порядка необходимо востановить в задаче распознавания радужной оболочки глаза~\cite{Matveev2010, Matveev2014, Bowyer2010}, в задаче описания трека частицы в адронном коллайдере~\cite{Dalila2018}.
Экспертная информация о кривой второго порядка отображает точки на плоскости в новое признаковое описание объекта. 
Каждая кривая аппроксимируется одной линейной моделью.
Модель, аппроксимирующая кривую, называется \textit{локальной моделью}.
Для аппроксимации всего контурного изображения требуется аппроксимация нескольких кривых второго порядка с использованием локальных моделей.
Вводятся следующие ограничения на изображения: а) изображение состоит только из кривых второго порядка; б) изображение аппроксимируется небольшим числом кривых второго порядка; в) известно количество и тип кривых на изображении.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{explanation}
  \caption {Пример: a) экспертная информация первого эксперта; b) исходные данные; c) экспертная информация второго эксперта}
  \label{intro:fig2}
\end{figure}

На рис.~\ref{intro:fig2} показан пример кривых второго порядка, а также экспертная информация о кривых. На рис.~\ref{intro:fig2} слева показана экспертная информация о первом эксперте. Используя эту информацию, первая кривая аппроксимируется линейной моделью, а вторая кривая представляет собой шум. На рис.~\ref{intro:fig2} справа показана экспертная информация второго эксперта. Используя эту информацию, вторая кривая аппроксимируется линейной моделью, а первая кривая представляет собой шум.

При аппроксимации нескольких кривых на одном контурном изображении строится мультимодель. Примером мультимоделей является случайный лес~\cite{Ishwaran2012}, бустинг деревьев~\cite{Tianqi2016}, смесь экспертов~\cite{Yuksel2012}.
В данной работе в качестве мультимодели рассматривается смесь экспертов.
Смесь экспертов — это мультимодель, которая линейно взвешивает локальные модели, которые аппроксимируют часть выборки.
Значения весовых коэффициентов зависят от объекта, для которого делается прогноз.
Для решения задачи используется вариационный EM--алгоритм~\cite{Dempster1977, Ebrahimpour2009, Peng1996, Grabovoy2021}.
Смесь экспертов имеет множество применений в ряде приложений.
В~\cite{Estabrooks2001} решается задача классификации текстов.
В~\cite{Cheung1995, Weigend2000, Cao2003, Mossavat2010, Sminchisescu2007, Tuerk2001, Yumlu2003} смесь экспертов используется для прогнозирования временных рядов, для распознавания речи, распознавания повседневной деятельности человека и прогнозирования стоимости ценных бумаг.
В~\cite{Ebrahimpour2009} рассматривалась смесь экспертов для решения задачи распознавания рукописных цифр на изображениях.

В качестве примера рассматривается задача аппроксимации изображения радужной оболочки глаза. На рис.~\ref{intro:fig1} показан пример изображения, которое необходимо аппроксимировать.
В данной работе рассматривается обработанное изображение в виде контурнуго изображения. Например рис.~\ref{intro:fig1} справа.
На рис.~\ref{intro:fig1} показаны две модели локальных кругов, которые соответствуют радужной оболочке глаза.

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{real_image}
\includegraphics[width=0.5\textwidth]{outline_image}
\caption{Пример изображения радужной оболочки глаза и его контурное представление. Слева: изображение радужной оболочки глаза. Справа: контурное изображение радужной оболочки и аппроксимирующие заданное изображение окружностей}
\label{intro:fig1}
\end{figure}

Для задачи аппроксимации радужной оболочки используется экспертная информация: радужная оболочка глаза аппроксимируется двумя концентрическими окружностями.
Экспертная информация используется для построения описания признаков точек плоскости, а также для построения регуляризатора в функции оптимизации.
Часть функции ошибок для оптимизации, использующая экспертную информацию, называется регуляризатором.
Таким образом, информация о том, что изображение являются окружностями задается видом признакового описания, а информация о том, что окружности концентрические задается с помощью специального регуляризатора.

Вычислительный эксперимент анализирует качество аппроксимации контурного изображения в зависимости от экспертной информации и уровня шума в синтетически сформированных данных. Проведен анализ качества аппроксимации радужной оболочки в зависимости от объема экспертной информации, которая использовалась для построения модели.
Каждое аппроксимированное изображение представляет собой отдельный набор точек, которые необходимо аппроксимировать.

\section{Постановка задачи востановления параметров кривых второго порядка на изображении}
Задано бинарное изображение:
\begin{gather}
\mathbf{M} \in \{0, 1 \}^{m_1\times m_2},
\end{gather}
где~$1$ соответствует черной точке изображения, а~$0$ соответствует белой точке фона.
Используя изображени~$\mathbf{M}$ строится выборка~$\mathbf{C}$, элементами которой являются координаты~$(x_i, y_i)$ черных точек:
\begin{gather}
\mathbf{C} \in \mathbb{R}^{N \times 2}.
\end{gather}
Каждый эксперт предполагает, что изображение состоит из кривой второго порядка~$\Omega$.
Пусть для множества точек~$\mathbf {C} \in \mathbb{R}^{N \times 2} $, образующих кривую $\Omega,$ задана экспертная информация о фигуре $E(\Omega)$.
Множество $E(\Omega)$ состоит из ожидаемого экспертом образа~$\Omega$ и множества его допустимых преобразований. На основе экспертного описания введем отображения в задачу аппроксимации:
\begin{gather}
\label{eq1}
	K_{x}\bigl(E(\Omega)\bigr): \mathbb{R}^{2} \rightarrow \mathbb{R}^{n}, \quad K_{y}\bigl(E(\Omega)\bigr): \mathbb{R}^{2} \rightarrow \mathbb{R},
\end{gather} 
где~$K_{x}$~--- отображение объектов в признаковое описание объектов,~$n$~--- число признаков, а ~$K_{y}$~---  отображение в целевую переменную для аппроксимации. Применяя отображения~$K_{x}, K_{y}$ для выборки~$\mathbf{C}$ поэлементно получаем:
\begin{gather}
\label{eq2}
	K_{x}\bigl(E(\Omega\bigr), \mathbf{c}) = \mathbf{x}, \quad  K_{y}\bigl(E(\Omega), \mathbf{c}\bigr) = y,
\end{gather}
где~$\mathbf{c} = (x_i, y_i)$ точка из множества точек $\mathbf{C}$.

Применяя отображения~\eqref{eq2} к исходному множеству точек~$\mathbf{C}$, получаем выборку
\begin{gather}
\label{eq4}
    \mathfrak{D} = \{(\mathbf{x}, y) \; | \; \forall \mathbf{c} \in \mathbf{C} \; \mathbf{x} = K_x(\mathbf{c}), \; y = K_y(\mathbf{c}) \}.
\end{gather}

Получаем, что исходная задача аппроксимации кривой~$\Omega$ сводится к аппроксимации выборки~$\mathfrak{D}$. В работе предполагается, что выборка~$\mathfrak{D}$ аппроксимируется линейной моделью:

\begin{gather}
	g(\mathbf{x}, \mathbf{w}) = \mathbf{x}^\mathsf{T} \mathbf{w},
\end{gather} 
где~$\mathbf{w}$~--- вектор параметров для аппроксимации.

Для нахождения оптимального вектора параметров~$\hat{\mathbf {w}}$ решается оптимизационная задача:
\begin{gather}
	\hat{\mathbf{w}} = \arg\min_{\mathbf{w}\in\mathbb{R}^n} \sum_{\left(\mathbf{x}, y\right) \in \mathfrak{D}}\|g(\mathbf{x}, \mathbf{w}) - y \|_2^2.
\end{gather}

Задача аппроксимации исходной кривой~$\Omega$ сводится к решению задачи линейной регрессии, т.е. к нахождению компонент вектора $\hat{\mathbf{w}}$.

В случае, когда на изображении~$K$ кривых второго порядка~$\Omega_1, \dots, \Omega_K$, где для каждой задана экспертная информация~$E_k = E(\Omega_k), \, k \in \{1, \dots, K\}.$ Получаем задачу построения мультимодели, которая называется смесью~$K$ экспертов.

\begin{definition}
Назовем мультимодель~$f$ смесью~$K$ экспертов
\begin{gather}
	f = \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V})g_k(\mathbf{w}_k),  \quad \pi_k(\mathbf{x}, \mathbf{V}): \mathbb{R}^{n\times |\mathbf{V}|} \rightarrow [0, \, 1], \quad \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V}) = 1, 
\end{gather}
где~$g_k$~--- локальная модель, называемая экспертом. Вектор~$\mathbf{x}$~--- признаковое описание объекта, $\pi_k$~--- шлюзовая функция, вектор $\mathbf{w}_k$~--- параметры локальной модели, $\mathbf{V}$ являются параметрами шлюзовой функции.
\end{definition}

Для каждой кривой второго порядка~$\Omega_k$ заданы отображения~\eqref{eq1}. Введем обозначения: $K_x^k\bigr(\mathbf{c}\bigr) = K_x\bigr(\Omega_k, \mathbf{c}\bigr)$ и $K_y^k\bigr(\mathbf{c}\bigr) = K_y\bigr(\Omega_k, \mathbf{c}\bigr)$. Используя локальные линейные моделей~$g_k$ строится мультимодель~$f$, описывающая кривые $\Omega_1, \dots, \Omega_K$ на изображении $\mathbf{M}$:
\begin{gather}
\label{5}
	f = \sum\limits_{\mathbf{c} \in \mathbf{C}} \sum_{k = 1}^{K} \pi_k(\mathbf{c}, \mathbf{V})g_k(K^k_{x}\bigl(\mathbf{c}), \mathbf{w}_k), 
\end{gather}
где~$\pi_k$~--- шлюзовая функция. В работе рассматривается случай, когда~$\mathbf{x}=K^1_{x}\bigl(\mathbf{c})=\cdots=K^K_{x}\bigl(\mathbf{ c}).$ В этом случае выражение~\eqref{5} переписывается в виде:
\begin{gather}
\label{5_1}
	f = \sum\limits_{\mathbf{c} \in \mathbf{C}} \sum_{k = 1}^{K} \pi_k(\mathbf{x}, \mathbf{V})g_k(\mathbf{x}, \mathbf{w}_k), 
\end{gather}
где шлюзовая функция~$\pi_k$ имеет вид:
\begin{gather}
\label{6}
	\pi_k(\mathbf{x}, \mathbf{V}): \mathbb{R}^{n\times |\mathbf{V}|} \rightarrow [0, \, 1], \; \; \; \; \sum\limits_{k = 1}^{K}\pi_k(\mathbf{x}, \mathbf{V}) = 1,
\end{gather}
где~$\mathbf{V}$~--- параметры функции шлюза, а $g_k$~--- локальная модель.

В работе рассматривается следующий вид шлюзовой функции:
\begin{gather}
    \boldsymbol{\pi}(\mathbf{x}, \mathbf{V}) = \text{softmax}\bigl(\mathbf{V}_1^{\mathsf{T}}\boldsymbol{\sigma}(\mathbf{V}_2^{\mathsf{T}}\mathbf{x}) \bigr),
\end{gather}
где~$\mathbf{V} = \{\mathbf{V}_1, \, \mathbf{V}_2\}$~--- параметры шлюзовой функции, $\mathbf{V}_1 \in \mathbb{R}^{p \times k}, \, \mathbf{V}_2 \in \mathbb{R}^{n \times p}$.

Для нахождения оптимальных параметров мультимодели решается оптимизационная задача:
\begin{gather}
\label{9}
\mathcal{L} = \sum\limits_{(\mathbf{x}, y) \in \mathfrak{D}} \sum\limits_{k = 1}^{K} \pi_k(\mathbf{x}, \mathbf{V})(y - \mathbf{w}_k^{\mathsf{T}}\mathbf{x})^2 + R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr) \rightarrow \min_{\mathbf{V}, \mathbf{W}},
\end{gather}
где~$\mathbf{W} = [\mathbf{w}_1, \dots, \mathbf{w}_k]$~--- параметры локальных моделей, $R\bigl(\mathbf{V}, \mathbf{W}, E(\Omega)\bigr)$~--- параметры регуляризации, основанные на экспертной информации.

\section{Построение признакового описания фигур}
\paragraph{Единое пространство для кривых второго порядка.} Произвольная кривая второго порядка, главная ось которой не параллельна оси ординат, задается  уравнением:
\[
\label{st:coef}
x^2 = B'xy+C'y^2+D'x+E'y+F',
\]
где на коэффициенты $B', C'$ накладываются ограничения, зависящие от типа кривой. Выражение~\eqref{eq2} для кривой второго порядка принимает вид:
\[
\label{st:K_map}
K_x\bigr(\mathbf{c}_i\bigr)=\left[x_iy_i, y_i^2, x_i, y_i, 1\right], \quad K_y\bigr(\mathbf{c}_i\bigr)=x_i^2,
\]
откуда получаем задачу линейной регрессии для восстановления параметров ~$B', C', D', E', F'$ по заданой выборке.

\paragraph{Окружность.} В качестве частного случая кривой второго порядка рассмотривается окружность.
Пусть~$(x_0, y_0)$~--- центр окружности, которую требуется востаноить на бинарном изображении~$\mathbf{M}$, а $r$~--- ее радиус.
Элементы выборки~$(x_i, y_i) \in \mathbf{C}$ представляют собой геометрическое место точек, которое аппроксимируется уравнением окружности:
\begin{gather}(2x_0)\cdot x_i + (2y_0)\cdot y_i + (r^2 - x_0^2 - y_0^2)\cdot 1 = x_i^2 + y_i^2. 
\end{gather}
Тогда выражение~\eqref{eq2} принимают следующий вид:
\begin{gather}
\label{10}
K_{x}(\mathbf{c}_i) = [x_i, \, y_i, \, 1] = \mathbf{x}, \,  K_{y}(\mathbf{c}_i) = x_i^2+y_i^2 = y.
\end{gather} 
Получаем задачу линейной регрессии~\eqref{eq4}.
Компоненты вектора~$\mathbf{w} = [w_0, \, w_1, \, w_2]^\mathsf{T}$, восстанавливают параметры окружности:
\begin{gather}
x_0 = \frac{w_0}{2}, \; y_0 = \frac{w_1}{2}, \; r = \sqrt{w_3 + x_0^2 + y_0 ^2}.
\end{gather}
